Flat histogram method comparison on 2D Ising Model

### 2020-04-21 ising/ising.tex
made some paper changes.

\begin{abstract}
  We examine the convergence of the Stochastic Approximation with a Dynamic
  update factor (SAD) algorithm applied to the 2D ising model. Comparison with
  SAMC and WL methods show that SAD performs robustly and without user input
  knowledge of an energy range. We confirm that SAD is more powerful in the
  common case in which the range of energies is not known in advance.
  \jpsays{production run WL is also tested and does well} 
\end{abstract}

@@

\jpsays{here might be a good place to mention production run WL}

@@

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can estimate the average deviation from the exact solution
using the relation~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control}:
\begin{align}
\braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
\end{align}
As per the implementation by Wang and
Landau~\cite{wang2001determining}, the total number of available energy
states are $N-1$.

\jpsays{add in the heat capacity formula here and comment out the average deviation form the exact solution for the entropy}

@@

We thank Klas Markstr\"{o}m for discussions regarding the exact computation of the partition function for the 2D Ising model. We also with to thank Johannes Zierenberg for helpful discussions regarding WL production simulations and
Martin Weigel for discussions regarding entropic population annealing.

>>>>

\begin{abstract}
  We examine the convergence of the Stochastic Approximation with a Dynamic
  update factor (SAD) algorithm applied to the 2D Ising model. Comparison with
  SAMC and WL methods show that SAD performs robustly and without user input
  knowledge of an energy range. We also compare SAD with a final production run WL simulation with fixed weights. While SAD is more powerful in the
  common case in which the range of energies is not known in advance, the 2D Ising model energy range is readily known and easily discovered by WL methods.
\end{abstract}

@@

The WL method can also be implemented as part of a production run which aids in
equilibration~\cite{gross2018massively}. WL is used to generate the weights
resulting in a flat-histogram~\cite{janke2017generalized}. Once a user-defined
minimum $\gamma$ is reached, the simulation switches to a final production run
with fixed weights. WL implemented in this way can ensure ergodicity and
detailed-balance. The converge of the production run is still impacted by the
choice of the minimum $\gamma$.

@@

\section{Ising Model}
The 2D Ising spin-lattice system is widely used as a testbed when
benchmarking or comparing Monte-Carlo
methods~\cite{ferdinand1969bounded, wang1999transition, trebst2004optimizing, barash2019estimating}. The 2nd order
phase transition behavior and the ability to directly calculate the
exact solution for finite lattices~\cite{beale1996exact, haggkvist2004computation} make the
system sufficiently interesting for such theoretical comparisons. It is
also important to note that direct comparison of the other methods can
be made with WL as its original implementation was done on this
system~\cite{wang2001determining,wang2001efficient}.
We test the convergence of several flat histogram methods
on the periodic 2D square lattice ferromagnetic Ising model with nearest
neighbor interactions~\cite{landau2004new}.
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j - h \sum_i s_i
\end{align}
The $N\times N$ spin system can take on values of $\sigma_i = \pm 1$
for up or down spins respectively. In the absence of a magnetic field ($h =
0$), We can write the Hamiltonian as follows~\cite{onsager1944crystal,
kaufman1949crystal}:
\begin{align}
\mathcal{H} = -\sum_{\braket{i,j}} \sigma_i \sigma_j
\end{align}
where the sum is over nearest neighbor spin sites. Beale showed that for finite
lattices the Density of States could directly be calculated from the partition
function~\cite{beale1996exact}
\begin{align}
Z = \sum_E g(E) e^{-{\beta E}}
\end{align}
% where $g(E)$ is the multiplicity of the system which is proportional to
% $D(E)$. We can estimate the average deviation from the exact solution
% using the relation~\cite{schneider2017convergence, shakirov2018convergence,
% barash2017control}:
% \begin{align}
% \braket{\epsilon(t)}_E = \frac{1}{N_E - 1} \sum_E \left( \frac{S(E,t) - S_{\text{Beale}(E)}}{S_{\text{Beale}(E)}} \right)
% \end{align}
where $g(E)$ is the multiplicity of the system which is proportional to
$D(E)$. We can estimate the maximum deviation in the canonical specific heat capacity $c_V$ from the exact solution~\cite{schneider2017convergence, shakirov2018convergence,
barash2017control,barash2017gpu}:
\begin{align}
\braket{\epsilon(t)}_E = \max_E \left( c_V(E,S,T,t) - c_V(E,S,T,t)_{\text{Beale}}\right)
\end{align}
% As per the implementation by Wang and
% Landau~\cite{wang2001determining}, the total number of available energy
% states are $N-1$.
Computing the specific heat capacity $c_V$ presents a difficult challenge for
any Monte-Carlo method due to fluctuations in the derivative of the internal energy.  Methods that accurately compute $c_V$ also by extension accurately compute the internal energy.

@@

We thank Klas Markstr\"{o}m for discussions regarding the exact computation of the partition function for the 2D Ising model. We also with to thank Johannes Zierenberg for helpful discussions regarding WL production run simulations and
Martin Weigel for discussions regarding entropic population annealing.
